---
title: "AML_Lab2"
author: "Josh Hyungyum Kim"
date: "2018 - 9 - 27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Build a hidden Markov model (HMM) for the scenario described above.

```{r, message=FALSE}

# Load packages
require(HMM)
require(entropy)

# 1.
# Build a HMM model

# trasition probability matrix
trans_mat <- matrix(0, ncol = 10, nrow = 10)
for(i in 1:9){
  trans_mat[i,i] <- 0.5
  trans_mat[i,i+1] <- 0.5
}
trans_mat[10,1] <- 0.5
trans_mat[10,10] <- 0.5

# emission probability matrix
emi_mat <- matrix(0, ncol = 10, nrow = 10)
for(i in 3:8){
  emi_mat[i,i] <- 0.2
  emi_mat[i,i+1] <- 0.2
  emi_mat[i,i+2] <- 0.2
  emi_mat[i,i-1] <- 0.2
  emi_mat[i,i-2] <- 0.2
}
emi_mat[1,c(1,2,3,9,10)] <- 0.2
emi_mat[2,c(1,2,3,4,10)] <- 0.2
emi_mat[9,c(1,7,8,9,10)] <- 0.2
emi_mat[10,c(1,2,8,9,10)] <- 0.2

# Build model with initHMM()
model1 <- initHMM(States = c(1,2,3,4,5,6,7,8,9,10),
                  Symbols = c(1,2,3,4,5,6,7,8,9,10),
                  startProbs = c(1,0,0,0,0,0,0,0,0,0),
                  transProbs = trans_mat,
                  emissionProbs = emi_mat)

model1
```

### 2. Simulate the HMM for 100 time steps.

```{r}

# 2. 
set.seed(12345)

# simulate 100 steps with simHMM()
res1 <- simHMM(model1, 100)
print(res1)
```

### 3. Discard the hidden states and compute the most probable path.

```{r}

# 3.
# Taking only observations
obs1 <- res1$observation

# Filtered and smoothed probs
filtered1 <- forward(model1, obs1)
smoothed1 <- posterior(model1, obs1)

# Most probable path
path1 <- viterbi(model1, obs1)
```


### 4. Compute accurace of the filtered and smoothed probability distributions and most probable path.

```{r}

# 4.

# Get states from the probability dists. by using apply
filter_states <- apply(prop.table(exp(filtered1)), MARGIN = 2, FUN=which.max)
smooth_states <- apply(prop.table(smoothed1), MARGIN = 2, FUN=which.max)

# True states
true_states <- res1$states

# Accuracy
list(Accuracy_of_filtered=table(true_states==filter_states),
     Accuracy_of_smoothed=table(true_states==smooth_states),
     Accuracy_of_path=table(true_states==path1))
```

### 5. Repeat wiht different samples.

```{r}

# 5.
set.seed(123456789)


# 3 more simulations 
res2 <- simHMM(model1, 100)
res3 <- simHMM(model1, 100)
res4 <- simHMM(model1, 100)

# Taking only observations
obs2 <- res2$observation
obs3 <- res3$observation
obs4 <- res4$observation

# Filtered and smoothed probs
filtered2 <- forward(model1, obs2)
smoothed2 <- posterior(model1, obs2)
filtered3 <- forward(model1, obs3)
smoothed3 <- posterior(model1, obs3)
filtered4 <- forward(model1, obs4)
smoothed4 <- posterior(model1, obs4)

# Most probable path
path2 <- viterbi(model1, obs2)
path3 <- viterbi(model1, obs3)
path4 <- viterbi(model1, obs4)

# Get states from the probability dists. by using apply
filter_states2 <- apply(prop.table(exp(filtered2)), MARGIN = 2, FUN=which.max)
smooth_states2 <- apply(prop.table(smoothed2), MARGIN = 2, FUN=which.max)
filter_states3 <- apply(prop.table(exp(filtered3)), MARGIN = 2, FUN=which.max)
smooth_states3 <- apply(prop.table(smoothed3), MARGIN = 2, FUN=which.max)
filter_states4 <- apply(prop.table(exp(filtered4)), MARGIN = 2, FUN=which.max)
smooth_states4 <- apply(prop.table(smoothed4), MARGIN = 2, FUN=which.max)

# True states
true_states2 <- res2$states
true_states3 <- res3$states
true_states4 <- res4$states

# Accuracy
list(Accuracy_of_filtered2=table(true_states==filter_states2),
     Accuracy_of_smoothed2=table(true_states==smooth_states2),
     Accuracy_of_path2=table(true_states==path2))

list(Accuracy_of_filtered3=table(true_states==filter_states3),
     Accuracy_of_smoothed3=table(true_states==smooth_states3),
     Accuracy_of_path3=table(true_states==path3))

list(Accuracy_of_filtered4=table(true_states==filter_states4),
     Accuracy_of_smoothed4=table(true_states==smooth_states4),
     Accuracy_of_path4=table(true_states==path4))

```

### 6. Is it true that the more observations you have the better you know where the robot is ?

```{r}

# 6. 

plot(apply(prop.table(exp(filtered1)), 2, entropy.empirical), col="darkblue",
     type="l", xlab="Steps", ylab="Entropy", main="Entropy plot of filtered distribution")
grid()
```
No

### 7. Consider any of the samples above of length 100. Compute the probabilities of the hidden states for the time step 101.

```{r}

# 7.

smoothed1[,100] %*% trans_mat
```

\clearpage

### 8. Appendix

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```

